Positivity-preserving adaptive Runge-Kutta methods: results

### 2020-01-27 latex/overview.tex
Update overview.tex
Cleaned up the text, added some graphics and tables.

TODOs: 
     Clean up section 'Implementation aspects of the Algorithm'
     Add the experimental results

@@

@@

Important things to point out:
Approach preserves linear invariants
Not cost anything if there are no negative components (exept for the test)

@@

@@

When computing the solution of an ODE $u ' = f(t,u) $ using a RK Method with $s$ Stages we first compute the stage values

@@

Afterwards we calculate the solution

@@

One reason why steps are requected is that the new solution does not comply with positifity. 
So it may be adventagious to adapt Equation\,\ref{eq:Combination} in such a way that $u^{n+1}$ is positive. This is done by adapting the weights $b$.
As we can see in Equation\,\ref{eq:Combination} $u^{n+1}$ depends linearly on the stage values.
When altering the weights the Order Conditions should be still meet. 
Because $A$ is already fixed the Order Conditions reduce to linear equations.
This means that finding the $b$ can be done by solving a linear Programming problem.

@@

For problems where only the combination of $f(t,u)$ and $u_0$ leads to positifity tests did not show promising results, additional it is not sure if the computed solution is reasonable.

@@

\subsection{Positifity}
For solving an PDE with the method of lines an spartial grid  with $m$ points is introduced. 
From this we get an ODE with an $u \in R^m$.

@@

has to be fulfilled.
This infers $p$ positivity constraints to the optimisation problem. These can be written as

@@

This causes unnecessary many inequality constraints on the LP-Problem. To simplify the problem we want to reduce the positifity constraints at places where there is no issue with positifity.

For this we write
\begin{equation}
u_i^n + \Delta b \sum_{j=0}^{s-1} b_j h_i^j  \geq 0   \;   \forall_{i \in h \subseteq \{1,\cdot,m \}} 
\end{equation}

Now we have to choose a $h \subseteq \{1,\cdots,m \}$ that include all indecies that are needed for determining the $b$.

A reasonable approach is to set $h_0 = \{ i \in \{1,\cdots,m \} |  u_i^{n+1}  < 0 \}$. 
After the solution with the reduced set is computed, the algorithm has to check if the positifity of the other values $i \notin h \subseteq \{1,\cdot,m \}$ is still meet. 
If the conditions is not meet a new set $h_{a+1} = \{ i \in \{1,\cdots,m \}|  u_i^{n+1}  < 0 \} \cup h_{a}$ using the new $u_i^{n+1}$ is generated and the LP-Problem is solved again. This process id repeated until a solution is found that leads to a $u_{n+1} \geq 0$. The choice of $h_{a+1}$ based on the $h_{a}$ ensures that the algorithm cannot get stuck in a loop. In the worst case $h_a$ is growing very slowly until $h_a = \{1,\cdots,m \}$.

@@

When enforcing a maximum value we can do the same. When both enforcing maximum and minimum values we have two separate sets of active constraints $k_{min}$ and $k_{max}$ where $k_{max}$ denotes the set of active maximum constraints and $k_{min}$ denotes the set of active minimum constraints. 
Here it is important to update these sets simultaneously.

\subsection{Order Conditions}

The Order Conditions are an linear equations system and can be written as $Q b = r$.

The vector $b \in R^s$ contains the b coefficients. $s$ is the number of stages. $Q$ is an $n \times s$ matrix, where $g$ is the number of Order Conditions. $r$ is an vector containing the right hand side of the Order Conditions

If the $A$ of a  preexisting RK-method is used, there is at least one solution for the linear equation system.
In order to get an optimization problem the equation system has to be underdetermined.
This implies that $rank{Q} < s$. 
Because the Quadrature Conditions are linear independent, there are at least $q$ independent vectors in $Q$. This leads to the $p < s$.
This implies that the number of stages has to be higher as the wanted required order.

@@

There are two possible ways to write down the adaptation of the $b$. The first way is to directly adapt the $b$ by adding a correction  $\Delta b$ that ensures positivity. 
This is the easiest way to adapt the $b$. An advantageous property is that the step size can be reduced after calculation the stage values by using the dense output order conditions.  
(Note: Recommendation for explicit methods)

The second approach is to choose $b$ as a convex combination of a set of $b$s. This approach is more useful if some embedded methods with certain properties are know. 
This approach is also shows more  predictable behavior because the resulting $u^{n+1}_b$ is also a convex combination of the solutions $u^{n+1}_{b_1},u^{n+1}_{b_2},\cdots$ 
(Note: Recommendation for implicit methods)

Both approaches are explained in more detail in the following sections.

\subsection{Direct Adaptation}

The Direct Adaptation approach leads to two set of constraints on the $b$.

@@

The weights $b = b_{orig} + \Delta b$ consist of the weights $b_{orig}$ of the original method and an adaption $\Delta b$. This $\Delta b \in ker \{ Q \}$.

To determine the optimal $b$ a objective function is needed.
Different types of objective functions have been tried. (Notebook: objectives.ipynb)

@@

\item[Quadrature]   The optimisation Problem opimises for $min(|bTq-r|)$
                            where q is the quadrature condition of the next higher order 
                            and r is the expected right hand side

@@

A reasonable property of the objective function would be that for a set of $b$ only constrained by the order condition the LP-Problem gives $b = b_{orig}$ as optimal solution.  $argmin(f_{optim}(b)) = b_{orig}$. In practice this also means that there is no need to solve solve the optimization Problem if no constraints are violated because the solution of this problem is already known. This is adventagious for performance reasons.
The different approaches have been tested for a test problem from Kopecz and Meister 2018 (Two substances reacting). The optimization problem was used at every step.
The magnitude of $b$ can get large if the Order Conditions are used as objective function.
By changing the $b$ the stability function is altered. This can lead to oscillations if the poles of the rhs are no longer in the stability region of the resulting method.
For this reason $||b-b_{orig}||$ seems like a good choice. The Norm $||x||_1$ can be implement the with Linear Programming using slack variables. We choose $||b_{optim}-b_{orig}||_1$ as objective function.

Large values of b can lead to numerical instability. Because $\sum_{i  = 1}^s b_i = 1$, a increase of $|b|$ leads to a combination of negative and positive values. Because the computation only is done with a finite precision (and usually in floating point) large $||b||$ can lead to absorption effects.

@@

If the original step $\Delta t$ was to large and there is no suitable $b$, we would like to calculate a $u^(n+1)$ using the stage values instead of rejecting the step. For this the time step is reduced after calculation the stage values. This can be done using the Order Conditions for dense Output (todo: Source for this).
We introduce $\theta \in [0,1]$. The new step taken has the length $\theta \Delta t$ whereas $\Delta t$ is the time step used when calculating the Stagevalues.
To calculate a reduced step the Order Conditions have to be adapted and a new $b_{orig}$ for the objective function (and more important the Error detection afterwards) has to be generated. 
(TODO: the $b_{orig}$ is still a open question)

@@

The dependency of $r(\theta)$ is nonlinear. This means that the maximum $\theta$ with a suitable $b$ can not be calculated by using a LP-Problem. This means that different $\theta$ have to be tested independently. 
(Note: we could include some algorithm to search for it)

\subsection{Convex combination of existing methods}
For some test problems (e.g. the Heat Equation solved with an implicit method), large distortions occur, even though a norm of the distortion of the b is used as a objective function. 
For implicit methods it is known that implicit euler steps lead to a positive solution for any $\Delta t$. The backward Euler is still only first Order but shows more exact results in certain cases than a method of higher Order that shows distortions of the solution caused by the larger error coefficients ($\Delta t \gg 0$, so they do not disappear because of the nature of the polynomial).
A possible method to fix this is to generate the new b based on a convex combination of embedded methods of RKM.

todo: write this down in more detail with proofs
$b$ is in a convex hull of al b
If all the used b complie with an order conditions the new method also has this order

@@

The optimisation problem is changed to:

$$u_{n+1}=u_n+\Delta K \vec{b} \geq 0$$

with

$$ \vec{b} = \sum_i^d a_i \vec{b_i}$$

$$ \forall_{i \in \{1, \cdots d \}} 0 \leq a_1 \leq 1$$

and the objective

@@

with some weighting factors $w_1,\cdots,w_d$. The weighting factors determine which is the preferred b to use.

Because the optimal solution of the unconstrained problem has to be the original $b$ the weight $w_0$ corresponding to $b_{orig}$ should be the smallest $w$. This weight is simply set to $w_{b_{orig}} = 0$. For the other weights $w_i = \frac{1}{Order\{b_i\}}$ is used.

For methods with $s \geq p$ we would still like to have some degrees of freedom for the highest order. For this we have to introduce at least one $b$ that complies with the order Condition of the highest order and $b \neq b_{orig}$.
These additional $b$ can be constructed by calculating a Span $S$ of the Kernel $ker Q$ of the Order Condition Matrix $Q$. For every vector $s \in S$ one set of weights with $b = b_{orig} + a s$ and one with $b = b_{orig} - a s$ is added. The $a$ is a parameter that determines how big a change is allowed.

@@

A Reduction of the step size suing the order Conditions for dense output would also be possible. For this 
Doing dense Output here is a bit complicated, but it would not help ensure positifity only (maybe) increase exactness. (Note: is it worth exploring this or is it better to just leave this open for further research it may be also interesting regardless the study of positifity)
Everything has downsides...

\subsection{Choice of Baseline method}

Existence of embedded methods and corresponding degrees of freedom for the b

\# TODO We could introduce some table here with the dimensions of the embedded methods for different methods and orders.

Stability region in comparison with the maximum timestep that ensures positifity.

Existence of a positive solution. This boils down to the question if there is a embedded first order method that ensures positivity. This is particularly interesting for Implicit methods, because the $\Delta t$ is not limited by the stability. 
If we have a embedded first Order Method we can ensure that there is always a positive solution, which might be of first Order.

\subsection{Error detection/Approximation}
The main question is: We see that for certain $b$ that comply with the Order Conditions we get results that do not resemble the solution. (some type of glitches, artefacts)

@@

\subsection{Stability region}

By adapting the $b$ we change the used RK Method. This also alters the stability function and therefore changes the region of absolute stability. 
It depends on the solved problem and the used $\Delta t$ on how many steps the stability function is altered. For Problems where only a small number of step are affected a bigger change in the stability function would be acceptable. For Problems with require a adaptation of the weights for every step one needs to make sure that the resulting method is stable.

@@

(Note:They do not behave like a Vector space because there is no scalar Product, and all the vector have to add up to 1... Is there something simmilar so we don not have to explain everything here?)

If the new $b$ is a chosen as a convex combination of $b$ it is easy to prove that the resulting  the Region of absolute stability is at least the Intersection of the region of absolute stability of the methods that define the convex combination.

If the $b$ is adapted directly there is no such proof.
A useful property of the resulting stability region is that it is similar to the stability region of the baseline method. 
(Note: is it possible to give some statement like: If the stability regions of the rkm used for multiple steps are similar, then the overall method is stable for the intersection of the stability regions)

Maybe place here some results on the stability region in dependence of the $b$
(Note: I have some proof handy that states that for certain $A$s the boundary of the stability region does only change in regard to a changing b in a limited way. But the proof is a bit convoluted. I have to look if there is a more compact way to prove it)

@@

(Note: It would also be possible to use another strategy for traversing all the combinations of $(p,\theta)$. Is it worth exploring these?)

@@

We want to show that $|R_{b_a}(z)|  < 1 ^ |R_{b_b}(z)| < 1\Rightarrow |R_{\chi b_a +(1- \chi) b_b}(z)| < 1$ with $\chi \in [0,1]$.
We write $\alpha = \chi$ and $\beta = 1-\chi$ for convenience.

@@

\subsection{Stability of direct adaptation}

@@

The stability function depending on the $b$ is expressed (Note: it is a bit complicated her because we cannot let it depend on the b directly, but only give b after a coordinate transformation, which also makes the expression easier because we can already introduce the order conditions here)
$m$ is the number of dimensions of the Kernel of the Order Condition $m = dim\{Ker Q \}$
The $b_1,\cdots,b_m$  are the Basis vectors of $Ker Q$. By this we can get the vector $w$ by $w = \left[b_1,\cdots,b_m\right]^{-1}(b-b_{orig})$.

@@

The border is defined by the implicit function 
\begin{equation}\label{eq:border}
|R(w,u+iv)|^2 -1 = 0
\end{equation}
This function can be vied either as $\mathbb{R}^m \times  \mathbb{C} \rightarrow \mathbb{R}$ or $\mathbb{R}^m \times  \mathbb{R}^2 \rightarrow \mathbb{R}$.
We are only concerned with the expansion or contraction. Because of this we are only interested in the change perpendicular to the Border of the stability region.

For a point $z_0= u_0 +i v_0 $ 
with $ |R(w,z_0)|^2 -1 = 0 $ the gradient of
 $|R(w,u,v)|^2$ is calculated. 
If the gradient does not vanish at $z = z_0$ Equation\,\ref{eq:border} Can be transformed in a function $\mathbb{R} \rightarrow \mathbb{R}$ depending on a single variable $a$ by setting $z = z_0 + n a$. The complex number $n$ denotes the normal vector $\vec{n} = \frac{\nabla |R(w,u,v))|^2}{\left| \nabla |R(w,u,v))|^2 \right|}$ of the border of the stability region written in complex notation. This defines the function $a(w)$ with $|R(w,z_0 + n a(w))|^2 -1 = 0$. 
This gives the condition that this approach is only applicable if the gradient does not vanish on the border of the stability function. This is not true in general but for the most common RKM.
The derivative $\frac{\mathrm d}{\mathrm d w} (a(w))$ can be calculated using the implicit function theorem. The defining function is $f: \mathbb{R} \rightarrow \mathbb{R}$ $f(a) = |R(b,z_0 + n a(w))|^2 -1 $ 
If explicit methods are used this function is known to be continuously differentiable because it can be written as a Polynomial of $w_1,\cdots,w_m,u$ and $v$. This is a necessary condition for the use of the of the Implicit function theorem. 
It is also known by definition, that for the point $w=0$ and $z = z_0$ the equation $ |R(w,z_0)|^2 -1 = 0 $ is satisfied.
For simplicity only the formulas for the derivative at the value $b = b_{orig}$ are given. This corresponds to $w_1 = \cdots = w_m = 0$.
The wanted derivative  is given by

@@

The Jacobian $J_{f,a}(w,a(w))$ evaluated at $w=0$ can be calculated using the directional derivative 
$$ J_{f,a}(b_{orig},a(w)) \Big|_{w=0} = 
 \frac{\partial f}{\partial a} (w,a(w)) \Big|_{w=0} = 
 \nabla |R(w,a(w)|^2 \vec{n} \Big|_{w=0}
= \left| \nabla|R(0,z_0)|^2 \right| = \left| \nabla|R_{b_{orig}}(z_0)|^2 \right|$$

>>>>

\usepackage{graphicx}
\usepackage{subcaption}

\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}
\usetikzlibrary{calc}
\usepackage{color}

\usepackage[backend=biber]{biblatex}
\addbibresource{pos_rk_stephan.bib}

@@

A big number of problems that preserve positifity reaction discretiesed PDEs...

@@

@@

Other Approaches: 
ssp104 methods: The idea is to construct a rkm Method out of explicit euler steps, that lead to a positive result for small enough $\Delta t$. (Todo: Read some literature regarding this and give a short explaination)

Implicit methods: Modified Patanka-Runge-Kutta (MPRK) \cite{kopecz_comparison_2019}

The idea of adapting the $b$ after calculation the stage values is used in \cite{ketcheson_spatially_2013}
In this case it is used to adapt the properties of the time integrator for a method of lines simulation on a PDE. With this approach it is possible to have different properties of the RKM at different parts of the domain.

@@

When computing the solution of an ODE $u ' = f(t,u) $ using a RK-Method with $s$ Stages the stage values $k^{(1)},\cdots,k^{(s)}$ are computed according to

@@

Based on these values the next solution $u^{n+1}$ is computed by

@@

One reason why steps are requected is that the new solution $u^{n+1}$ does not comply with positifity. 
So it may be adventagious to adapt Equation\,\ref{eq:Combination} in such a way that $u^{n+1}$ is positive. This is done by adapting the weights $b$.
As it can be see in Equation\,\ref{eq:Combination} $u^{n+1}$ is a linear combination of the stage values and the previous solution $u^n$.
When altering the weights the properties of the RKM should be preserved. Especially the Order of the RKM.  
To ensure this the Order Conditions should be still meet. 
Because the coeffeicents $a_{jl}$ are already fixed the Order Conditions reduce to linear equations.
This means that finding the $b$ can be done by solving a linear Programming problem.

This approach to ensure positifity preserves linear invariants like mass conservation, because it is a regular RKM. 
If there are no negative values in $u^{n+1}$ with the original $b$, then the method does not cause additional effort, because it only has to solve the LP-Problem if the regular RKM does not lead to a positive $u^{n+1}$. (exept for the test)
No additional evaluations of the rhs.

@@

For problems where only the combination of $f(t,u)$ and $u_0$ leads to the positifity of the exact solution, tests did not show promising results. Additional it is not certain if the computed solutions would be reasonable.

@@

\subsection{Positifity}

For the further discussion we are concerned with a System of $m$ coupled ODEs. 
In this setup  $u \in \mathbb{R}^m$.
Systems of ODEs with a high number of dimensions can appear when solving a PDE with the method of lines.

@@

has to be fulfilled.
This infers $m$ positivity constraints to the optimisation problem. These can be written as

@@

This causes unnecessary many inequality constraints on the LP-Problem. To simplify the problem one want to reduce the number of positifity constraints by ignoring it for ODEs where there is no issue with positifity.

The new positifity constraints can be formulated as
\begin{equation}
u_i^n + \Delta b \sum_{j=0}^{s-1} b_j h_i^j  \geq 0   \;   \forall_{i \in h \subseteq \{1,\cdot,m \}} 
\end{equation}

For this a set $h \subseteq \{1,\cdots,m \}$ that include all indecies that are needed for determining the $b$ has to be chosen.

A reasonable approach is to set $h_0 = \{ i \in \{1,\cdots,m \} |  u_i^{n+1}  < 0 \}$. 
After the new $b$ and $u^{n+1}$ is computed using $h_0$, the algorithm has to check if the positifity of the other values $i \notin h_0 \subseteq \{1,\cdots,m \}$ is still meet. 
If the conditions is not meet a new set $h_{a+1} = \{ i \in \{1,\cdots,m \}|  u_i^{n+1}  < 0 \} \cup h_{a}$ using the new $u^{n+1}$ is generated and the LP-Problem is solved again. This process id repeated until a solution is found that leads to a $u^{n+1} \geq 0$. The choice of $h_{a+1}$ based on the $h_{a}$ ensures that the algorithm cannot get stuck in a loop. In the worst case $h_a$ is growing very slowly until $h_a = \{1,\cdots,m \}$.

@@

When enforcing a maximum value the number of constraints can be reduced using the same technique. When both enforcing maximum and minimum values two separate sets of active constraints $k_{min}$ and $k_{max}$ are used. The set $k_{max}$ denotes the active maximum constraints and the set $k_{min}$ denotes the active minimum constraints. 
Here it is important to update these sets simultaneously.

\subsection{Order Conditions}\label{sec:OrderCond}

The Order Conditions for a RKM are a set of Equations depending on $a_{jl}$, $c_1,\cdots,c_s$ and $b_1,\cdots,b_s$. In these Equations $b_1,\cdots,b_s$ only appear linearly \cite{hairer_runge-kutta_1993}.
If $a_{jl}$ and $c_1,\cdots,c_s$ are already known, the Order Conditions for the Order $p$ are an linear equations system and can be written as $Q b = r$.

The vector $b \in \mathbb{R}^s$ contains the weights and $Q$ is an $n \times s$ matrix, where $g$ is the number of Order Conditions. The vector $r$ contains the right hand side of the Order Conditions.

If the $A$ of a  preexisting RK-method is used, there is at least one solution for the linear equation system.
In order to get an optimization problem the equation system has to be underdetermined.
This implies that $\mathrm{rank}(Q) < s$. 
Because the Quadrature Conditions are linear independent, there are at least $p$ independent vectors in $Q$. This leads to the condition $p < s$.
This implies that the number of stages has to be higher as the wanted order.

@@

There are two possible ways to write down the adaptation of the $b$. The first way is to directly adapt the original weights $b_{orig}$ by adding a correction $\Delta b$ that ensures positivity.
This is the easiest way to adapt the $b$. An advantageous property is that the step size can be reduced after calculation the stage values by using the dense output order conditions.  
(Note: Recommendation for explicit methods)
The direct adaption approach is explained in more detail in section\,\ref{sec:direct}.

The second approach is to choose $b$ as a convex combination of a set of $b$s. This approach is more useful if some embedded methods with certain properties are know. 
This approach also shows more  predictable behavior because the resulting $u^{n+1}_b$ is also a convex combination of the solutions $u^{n+1}_{b_a},u^{n+1}_{b_b},\cdots$ 
(Note: Recommendation for implicit methods)
The direct adaption approach is explained in more detail in section\,\ref{sec:convex}

Both approaches are shown in figure\,\ref{fig:b_space}.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_direct}
        \caption{Direct Adaptation}
        \label{fig:b_direct}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_convex}
        \caption{Convex Combiantion}
        \label{fig:b_convex}
    \end{subfigure}
    \caption{Graphical representation of the two different approaches to adapt the weights}\label{fig:b_space}
\end{figure}

\subsection{Direct Adaptation}\label{sec:direct}

When directly adapting the $b$ a new $b$ that complies with the Order Conditions is chosen.   
The weights $b = b_{orig} + \Delta b$ consist of the weights of the original method $b_{orig}$ and an adaption $\Delta b$. This $\Delta b \in \mathrm{ker}(Q)$. 
In Figure\,\ref{fig:b_direct} this is is drawn. The red line represents the subspace of $b$ that compile with the Order Conditions. Both $b_{orig}$ and $b$ are in this subspace.

To run it as a LP-Problem a set of constraints is constructed.  
Also a objective function is needed to determine the optimal $b$.

\subsubsection{Constraints}
The Direct Adaptation approach leads to two set of constraints on the $b$.

@@

\subsubsection{Objective Function}
Different types of objective functions have been tried.

@@

\item[Quadrature]   The optimisation Problem opimises for $min(|b^Tq-r|)$
                            where q is the quadrature condition of the next higher order 
                            and r is the expected right hand side

@@

A reasonable property of the objective function would be that for the LP-Problem only constrained by the order condition the optimal solution is $b = b_{orig}$. 
This can be achieved by $argmin(f_{optim}(b)) = b_{orig}$. 
In practice this also means that there is no need to solve solve the optimization Problem if no constraints are violated because the solution of this problem is already known. This is adventagious for performance reasons.
The different approaches have been tested on a linear test problem from \cite{kopecz_unconditionally_2018}. The optimization problem was used at every step.
The magnitude of $b$ can get large if the Order Conditions are used as objective function.
Large values of b can lead to numerical instability. Because $\sum_{i  = 1}^s b_i = 1$, a increase of $||b||$ leads to a combination of negative and positive values. Because the computation only is done with a finite precision (and usually in floating point) large $||b||$ can lead to absorption effects.    
By changing the $b$ the stability function is altered. This can lead to oscillations if the poles of the RHS are no longer in the stability region of the resulting method.
For these reasons $||b-b_{orig}||$ seems like a good choice. The 1-Norm can be implement the with Linear Programming using slack variables. We choose $||b_{optim}-b_{orig}||_1$ as objective function.

@@

If the original step $\Delta t$ was to large and there is no suitable $b$, we would like to calculate a $u^{n+1}$ using the stage values instead of rejecting the step. For this the time step is reduced after calculation the stage values. This can be done using the Order Conditions for dense Output. These can be found in \cite{hairer_runge-kutta_1993}.
We introduce $\theta \in (0,1]$. The new step taken has the length $\theta \Delta t$ whereas $\Delta t$ is the time step used when calculating the Stagevalues.
To calculate a reduced step the Order Conditions have to be adapted and a new $b_{orig}$ for the objective function (and more important the Error detection afterwards) has to be generated. 
(TODO: the $b_{orig}$ is still a open question)

@@

The dependency of $r(\theta)$ is nonlinear. This means that the maximum $\theta$ with a suitable $b$ can not be calculated by solving a LP-Problem. This means that different $\theta$ have to be tested independently. 
(Note: we could include some algorithm to search for it)

\subsection{Convex combination of existing methods}\label{sec:convex}
For some test problems (e.g. the Heat Equation solved with an implicit method), large distortions occur, even though a norm of the distortion of the b is used as a objective function. 
For implicit methods it is known that implicit euler steps lead to a positive solution for any $\Delta t$. The backward Euler is still only first Order but shows more exact results in certain cases than a method of higher Order that shows distortions of the solution caused by the larger error coefficients ($\Delta t \gg 0$, so they do not disappear because of the nature of the polynomial).
A possible method to fix this is to generate the new b based on a convex combination of embedded methods of RKM.

\subsubsection{Constraints}
At first a set of embedded methods of the RKM $b_1,\cdots,b_d$ are needed. 
If all the used b complies with an order conditions the new method also has this order.
This can is straightforward using

@@

The new $b$ is defined as 
\begin{equation}\label{eq:b_convex_def}
\vec{b} = \sum_i^d a_i \vec{b_i} = Ba
\end{equation}

where $B=\left[b_1,\cdots,b_d \right]$ is a matrix containing the embedded methods.  
To ensure the that Equation\,\ref{eq:b_convex_def} yields a convex combination following constraints are enforced for the  the parameters $a,\cdots,a_d$.

\begin{equation}
 0 \leq a_1 \leq 1  \forall_{i \in \{1, \cdots d \}}
\end{equation}
\begin{equation}
 \sum_{i=1}^d a = 1
\end{equation}

The positifity constraint changes to

$$u_{n+1}=u_n+\Delta K B a \geq 0$$

\subsubsection{Objective Function}

As an Objective function the function

@@

is used.  The $w_1,\cdots,w_d$ are weighting factors that determine which is the preferred $b_i$ to use.

Because the optimal solution of the unconstrained problem has to be the original $b$ the weight $w_0$ corresponding to $b_{orig}$ should be the smallest $w$. This weight is simply set to $w_{b_{orig}} = 0$. For the other weights $w_i = \frac{1}{\mathrm{Order} \{b_i\}}$ is used.

\subsubsection{Construction of Embedded Methods}
For methods with $s \geq p$ it is useful to have some degrees of freedom for the highest order. For this at least one $b$ that complies with the order Condition of the highest order and $b \neq b_{orig}$ has been introduced.
These additional $b$ can be constructed by calculating a Span $S$ of the Kernel $\mathrm{ker} (Q)$ of the Order Condition Matrix $Q$. For every vector $s \in S$ one set of weights with $b = b_{orig} + a s$ and one with $b = b_{orig} - a s$ is added. The $a$ is a parameter that determines how big a change is allowed.

@@

A Reduction of the step size suing the order Conditions for dense output would also be possible. For this a set of embedded methods $b_1(\Theta),\cdots,b_d(\Theta)$ has to be constructed. This was not done in this research.

(Note: Doing dense Output here is a bit complicated, but it would not help ensure positifity only (maybe) increase exactness.  is it worth exploring this or is it better to just leave this open for further research it may be also interesting regardless the study of positifity)
Everything has downsides...

\subsection{Choice of Baseline Method}
An important property for the used baseline method is the Existence of embedded methods and the degrees of freedom for the b.
As noted in Section\,\ref{sec:OrderCond} the number of Stages has to be higher than the Order. 
Another important property is the number of degrees of freedom for the choice of the new weights. 
These can be calculated using $s-\mathrm{rank}(Q)$%$\mathrm{dim}(\mathrm{ker}(Q))$.
The number of degrees of freedom for different methods are shown in table\,\ref{table:DOF_exp} for explicit methods and in table\,\ref{table:DOF_imp} for implic methods.

\begin{table}[h!]
\centering    %Generated below============ 
\begin{tabular}{|l |c c c c c c |} 
 \hline 
Order &1&2&3&4&5&6 \\ 
 \hline Classical RK4&3&2&0&0& - & -  \\ 
 SSPRK(10,4)&9&8&6&4& - & -  \\ 
 Cash-Karp RK5(4)6&5&4&2&1&0& -  \\ 
 Dormand-Prince RK5(4)7&6&5&3&1&0& -  \\ 
 \hline 
 \end{tabular}
 \caption{Degrees of Freedom for Choice of $b$} %Generated above============ 
 \label{table:DOF_exp}
 \end{table}

 \begin{table}[h!]
\centering   %Generated below============ 
 \begin{tabular}{|l |c c c c c c |} 
 \hline 
Order &1&2&3&4&5&6 \\ 
 \hline Implicit Euler&0& - & - & - & - & -  \\ 
 Im-Euler 2&2&1& - & - & - & -  \\ 
 Im-Euler 3&5&4&2& - & - & -  \\ 
 Im-Euler 4&9&8&6&3& - & -  \\ 
 \hline 
 \end{tabular}
 \caption{Degrees of Freedom for Choice of $b$} %Generated above============ 
 \label{table:DOF_imp}
 \end{table}

For all methods with the number of stages equal to the order of the method, there are no degrees of freedom without reducing the order. 
If the Classical RK4 method is used the Order has to reduced more because the RK4 method does not have embedded methods of order 3.
Contrary to this methods with $s > p$ have degrees of freedom even without reducing the order.

Another important property is the stability region in comparison with the maximum $\Delta t$ that ensures positifity. For this approach to be useful the RK-method has to yield negative results for $\Delta t$ smaller than the biggest stable $?Delta t$.

(Note: up to this point i do not understand why this is happening. We only see it occur depending on the Proble-RKM combination)

A third property is the existence of a positive solution. This corresponds to the question if there is a embedded first order method that ensures positivity. This is particularly interesting for Implicit methods, because the $\Delta t$ is not limited by the stability. 
If we have a embedded first Order Method we can ensure that there is always a positive solution, which might be of first Order. (Todo: Source)

\subsection{Error detection/Approximation}
(Todo: edit this whole section, Add some mathematically more rigorous statement)

The main question is: We see that for certain $b$ that comply with the Order Conditions we get results that do not resemble the solution. (some type of glitches, artefacts)

@@

\subsection{Stability Region}

By adapting the $b$ the used RK Method is changed. This also alters the stability function and therefore changes the region of absolute stability. 
It depends on the solved problem, the RKM and the used $\Delta t$ on how many steps the stability function is altered. For Problems where only a small number of step are affected a bigger change in the stability function would be acceptable. For Problems with require a adaptation of the weights for every step one needs to make sure that the resulting method is stable.

@@

\subsubsection{Convex Combination}
If the new $b$ is a chosen as a convex combination of $b$ it is easy to prove that the resulting  the Region of absolute stability is at least the Intersection of the region of absolute stability of the methods that define the convex combination. (Compare Section\,\ref{proof:convex_comb}). This also means that if all the used embedded methods used to construct the new $b$ are A-Stable, the resulting method is also A-Stable.

\subsection{Direct Adaptation}
If the $b$ is adapted directly there is no such proof.
A useful property of the resulting stability region is that it is similar to the stability region of the baseline method. 
(Note: is it possible to give some statement like: If the stability regions of the rkm used for multiple steps are similar, then the overall method is stable for the intersection of the stability regions)

In Section\,\ref{sec:deriv_sr_b} a proof is given that for most methods a small change of $b$ does not alter the stability function dramatically.

@@

(Note: It would also be possible to use another strategy for traversing all the combinations of $(p,\theta)$. Is it worth exploring these? Eg, a visect algorithm for $\Theta$ or linear approximations)

@@

We want to show that $(|R_{b_a}(z)|  < 1) (|R_{b_b}(z)| < 1) \Rightarrow |R_{\chi b_a +(1- \chi) b_b}(z)| < 1$ with $\chi \in [0,1]$.
We write $\alpha = \chi$ and $\beta = 1-\chi$ for convenience.

@@

\subsection{Stability of direct adaptation}\label{sec:deriv_sr_b}

@@

The stability function depending on the $b$ is expressed (Note: it is a bit complicated her because we cannot let it depend on the b directly, but only give b after a coordinate transformation, which also makes the expression easier because we can already introduce the order conditions here)
$m$ is the number of dimensions of the Kernel of the Order Condition $m = \mathrm{dim}\{\mathrm{Ker} (Q) \}$
The $b_1,\cdots,b_m$  are the Basis vectors of $\mathrm{Ker} (Q)$. By this we can get the vector $w$ by $w = \left[b_1,\cdots,b_m\right]^{-1}(b-b_{orig})$.

@@

The border is defined by the implicit function 
\begin{equation}\label{eq:border}
|R(w,u+iv)|^2 -1 = 0
\end{equation}
This function can be viewed either as $\mathbb{R}^m \times  \mathbb{C} \rightarrow \mathbb{R}$ or $\mathbb{R}^m \times  \mathbb{R}^2 \rightarrow \mathbb{R}$.
We are only concerned with the expansion or contraction. Because of this we are only interested in the change perpendicular to the Border of the stability region.

For a point $z_0= u_0 +i v_0 $ 
with $ |R(w,z_0)|^2 -1 = 0 $ the gradient of
 $|R(w,u,v)|^2$ is calculated. 
If the gradient does not vanish at $z = z_0$ Equation\,\ref{eq:border} Can be transformed in a function $\mathbb{R}^m \times \mathbb{R} \rightarrow \mathbb{R}$ depending on $w$ and a single variable $a$ by setting $z = z_0 + n a$. The complex number $n$ denotes the normal vector $\vec{n} = \frac{\nabla |R(w,u,v))|^2}{\left| \nabla |R(w,u,v))|^2 \right|}$ of the border of the stability region written in complex notation. This defines the function $a(w)$ with $|R(w,z_0 + n a(w))|^2 -1 = 0$. 
This implies that this approach is only applicable if the gradient does not vanish on the border of the stability function. This is not true in general but for the most common RKM.
The derivative $\frac{\mathrm d}{\mathrm d w} (a(w))$ can be calculated using the implicit function theorem. The defining function is 
\begin{equation}\label{eq:f(w,a)}
f: \mathbb{R}^m \times \mathbb{R} \rightarrow \mathbb{R} f(a) = |R(w,z_0 + n a)|^2 -1 
\end{equation}

If explicit methods are used this function is known to be continuously differentiable because it can be written as a Polynomial of $w_1,\cdots,w_m,u$ and $v$. This is a necessary condition for the use of the of the Implicit function theorem. 
It is also known by definition, that for the point $w=0$ and $z = z_0$ the equation $ |R(w,z_0)|^2 -1 = 0 $ is satisfied.
For simplicity only the formulas for the derivative at the value $b = b_{orig}$ are given. This corresponds to $w_1 = \cdots = w_m = 0$.
The wanted derivative  is given by

@@

The Jacobian of equation\,\ref{eq:f(w,a)} $J_{f,a}(w,a(w))$ evaluated at $w=0$ can be calculated using the directional derivative
\begin{align*}
 J_{f,a}(b_{orig},a(w)) \Big|_{w=0} =& 
 \frac{\partial f}{\partial a} (w,a(w)) \Big|_{w=0} = 
 \nabla |R(w,a(w)|^2 \vec{n} \Big|_{w=0} \\
=& \left| \nabla|R(0,z_0)|^2 \right| = \left| \nabla|R_{b_{orig}}(z_0)|^2 \right|
\end{align*}
